State of the Practice for Lattice Boltzmann Method Software

### 2021-12-22 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updated paper to use the new better aspect ratio figures
Updates up to and including section 4

\section{Introduction}

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice} such as
reflective and non-reflective boundaries. They consider the behaviour of a
collection of particles as a single unit at the mesoscopic scale, between the
nanoscopic and microscopic scales. LBM solvers predict the positional
probability of a collection of particles moving through a lattice structure
following a two step process: i) streaming, where the particles
move along the lattice via links; and, ii) colliding, where energy and
momentum is transferred among particles that collide \citep{bao2011lattice}. As
an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds numbers.
\wss{Zahra might have a better example?  A journal publication isn't likely
going to accept a figure produced for another paper.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development to provide insight into
its best practices and to offer guidance for future development.

@@

The full template consists of 108 questions categorized under 9 qualities:
\begin{inparaenum}
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum} 
The questions were designed to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. The measures are grouped under headings for
each quality, and one for summary information
(Figure~\ref{measurement_template_image}).   The summary section provides
general information, such as the software name, number of developers, etc.  We
follow the definitions given by \citet{GewaltigAndCannon2012} for the software
categories.  Public means software intended for public use.  Private means
software aimed only at a specific group, while the concept category is used for
software written simply to demonstrate algorithms or concepts. The three
categories of development models are: open source, where source code is freely
available under an open source license; free-ware, where a binary or executable
is provided for free; and, commercial, where the user must pay for the software
product.

@@

We requested interviews from all 24 packages, except Munubi, since Munubi was
added after our ethics approved study window had expired.  We were able to
recruit 4 developers to participate in our study.  To send out interview
requests we found we found contacts on the projects’ website, or code
repository, or publications, or the biographic pages of the teams’ institutions.
We send at most two interview request emails to a contact for each software
package.  Meeting will typically be held using on-line meeting software, like
Zoom or Teams.  This facilitates recording and automatic transcription of the
meetings.

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\hline
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\hline
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\hline
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
			Reflexive Boundary Condition (Yes or No), MFl for Multifluid (Yes or
			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/correctnessverifiability_chart}
		\caption{AHP Surface Correctness and Verifiability Score}
		\label{Fig_CorrectnessVerifiability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/reliability_chart}
		\caption{AHP Surface Reliability Score}
		\label{Fig_Reliability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/robustness_chart}
		\caption{AHP Surface Robustness Score}
		\label{Fig_Robustness}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/usability_chart}
		\caption{AHP Surface Usability Score}
		\label{Fig_Usability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/maintainability_chart}
		\caption{AHP Maintainability Score}
		\label{Fig_Maintainability}
	\end{center}
\end{figure}

@@

We measured the total number of source code files for each project. A larger
number of source files is associated with increased reusability, due to our
assumption that this indicates increased modularization. Some packages have more
features than others.  This is assumed to contribute to reusablility, since they have
more source code for potential reuse. The software packages were also reviewed
for the presence of API documentation, which indicates that a software package
was developed with interaction between other software applications in mind.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/reusability_chart}
		\caption{AHP Reusability Score}
		\label{Fig_Reusabilty}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/understandability_chart}
		\caption{AHP Understandability Score}
		\label{Fig_Understandability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/visibilitytransparency_chart}
		\caption{AHP Visibility and Transparency Score}
		\label{Fig_VisibilityTransparency}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\centering
		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart}
		\caption{AHP Overall Score}
		\label{Fig_OverallScore}
\end{figure}

@@

@@

Interviews with developers confirmed that these software packages are developed
by domain experts with backgrounds in physics, mathematics and mechanical
engineering. \wss{Could point to surveys that have shown the same thing.}  It
was noted in these interviews that some of the developers do not have formal
software engineering education. Some of the development teams include computer
scientists. Despite a lack of visible domain documentation and a resulting lower
surface correctness and verifiability score, it is clear that some of the
software packages were developed by teams with significant domain knowledge on
account of the academic backgrounds of their developers.

@@

Several developers alluded to difficulty with testing the correctness of large
numbers of features, and some even manually tested program output. The use of
well defined unit testing tools could decrease the time spent testing some
feature.  \wss{link this to recommendations}

@@

Interviews with developers revealed several usability issues. Some users have
misunderstood the boundaries of LBM and CFD, and have combined or applied
methods that are not physically sound. Sometimes users have applied LBM to
poorly defined or inappropriate fluid dynamics problems. For example, they may
wish to model flow through or around a structure despite having limited
information about the structure or its environment, and having little previous
knowledge of CFD. The users do not realize the limitations of the methods, of
the software, and do not understand the requirements to properly model a problem
with the software. As the developer of TCLB noted, such software packages are
not designed to be used ``out of the box'' in a plug and play fashion, and it
could take months or more to set up the CFD problems correctly. Developers of
some software packages, including ESPResSo, mitigated this by editing the source
code to prevent users from ``combining methods that are not physically sound
together'', and by updating the documentation to better inform users of LBM
limitations, and of the requirements to properly model appropriate problems,
including what algorithms and parameters to use.

Some additional but infrequent software usability issues were commented on by
the developers. Users have had trouble with installation and understanding how
to maneuver the interfaces and how to set up or run models. These issues are
addressed by various user support models, including frequently asked questions
sections on the software package websites, user guides, and hardware and
software requirements specifications.

One software package (ESPResSo) changed some of its scripting language to Python
to make it more usable. The developer commented that this was ``the biggest step
in terms of usability over the years'', further commenting that ``most people in
the field know [Python]'' and that ``it's easy to learn''.

@@

Interviews with developers revealed that most projects do have a defined process
for accepting contributions from team members. The packages rarely get
contributions from outside developers, but the process would be similar as for
the aforementioned group.

Contributions are made through GitHub, and are then reviewed and pulled by lead
developers, often with consultation with a group of core developers depending on
the organizational model. Continuous integration is part of the process for some
packages.

Some developers noted that their software package does not have well defined
contributing guide in the repository, but it might be a good idea to add one in
the near future. They would be happy to see contribution from outside of their
organization, but currently this does not happen.

Furthermore, maintainability has been addressed by increasing source code
modularity, reducing duplicate information, and improving abstraction by
developing well defined interfaces. This was noted by the developers of ESPResSo
and pyLBM. Several software packages have had sections of their code base
redeveloped with languages that the developers felt are more understandable and
readable, and that are better supported, such as Python. Data structures have
also been redeveloped and storage has been improved. A developer of pyLBM
mentioned that the geometries and models of their system had been ``decoupled'',
using abstraction and modularization of the source code, to make it ``very easy
to add [new] features''.

@@

Software packages were not quantitatively measured for modifiability. In this
project we asked developers to comment on modifiability when we interviewed
them. Specifically, we asked if ease of future changes to the system, modules,
and code blocks was considered when designing the software. We also asked if any
measures had been taken to ensure the ease of future changes. All of the
developers that were interviewed noted that the ease of future changes was
considered and that measures to ensure it had been taken, including requiring
the separation of software components in the source code architecture.

A high degree of code modularity and abstraction was noted by developers as a
measure to ensure the ease of future changes. This can be ensured by separating
components and hiding information behind well defined interfaces. The developer
of ESPResSo also noted that some of the code base was transitioned from C to
C++, which could ease modifiability of that software package. The developer of
TCLB noted that their software package was designed to allow for the addition of
some LBM features, but changes to major aspects of the system would be
difficult. For example, ``implementing a new model will be an easy
contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
Furthermore, the package was designed with flexible data structures and storage
in mind.

Some software packages, like Palabos, provide validation benchmarks for their
core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
intent of these benchmarks is to showcase the validity and usefulness of the
package to stimulate the development of third-user extensions. The Palabos
package identifies as a development framework for modeling problems in various
CFD areas.

@@

Software developers noted that they believe users have generally found their
packages to be understandable. The interviewed developer of ESPResSo commented
that some users have attempted to run physically incompatible LBM methods, and
the solution was to edit the code to prevent such combinations, as well as to
update the documentation to prevent misunderstanding the methods. Similarly, a
developer of pyLBM noted that some users had issues setting up parameters for
LBM schemes. The solution to this was to update the interface where these
parameters are set, as well as to add functionality to test the stability of the
parameters. A developer of OpenLB noted that some users lack the background
knowledge to easily model fluid dynamics problems using their software. A
frequently asked questions section was added to their package website to help
users find answers to common questions. The package also has detailed
documentation, including guides and usage requirements specification, to better
help users understand the software.

@@

The interviewed developer of ESPResSo noted that all major additions to their
package had accompanying changes to artifacts and documentation. They noted that
considerable effort had been put into the documentation. They further commented
that they want to lower the entry barrier for new developers, and because of
that their package has a considerable amount of developer documentation. This
documentation informs developers on how to get started, and orients them to the
artifacts, source code, and system architecture, as well as how the software
package build system works, and how the coupling between the simulation engine
and the interface works.

Developers noted the importance of documentation for both the users and
developers of their software. New features are always added to the
documentation. The developers use documentation to stay up to date on the status
of the software package, and to help expand features, like computational models
or algorithms. This is necessary so that the coding standard for these models is
kept consistent with new developers.

The importance of documentation for both users and developers was stressed
throughout the interviews. However, it was noted several times that a lack of
time and funding has a negative affect on the documentation. Most of the
developers are scientific researchers evaluated on the scientific papers that
they produce. Writing and updating documentation is something that is done in
their free time, if that time arises. Sometimes it is a last priority for the
developers. Finding ways to hasten updating documentation would increase the
frequency of such updates and benefit both users and developers.

The developer of OpenLB noted the use of documentation generators like Doxygen.
It would be advisable for more projects to use such automatic document
generation tools, since some projects do not do this.

Developers were asked to comment on the obstacles in their development process.
The developer of ESPResSo noted that a lot of their source code had been written
with a specific application in mind, and that there is too much coupling between
components. Addressing this issue would help with code modifiability and
reusability. Updating the development process would help resolve this issue and
prevent such issues in the future. Improving the visibility of software changes
and a peer review process would also help. Improving the software engineering
education or experience of developers is also an idea that was brought up by
several developers. Specifically, developers should always write code that is
decoupled and modular, and should keep in mind the visibility of their
contributions by better updating documentation and ensuring that their
contributions are transparent to the rest of the developers. This would help
catch issues in the contributions, and improve source code maintenance.

According to the interviewed developer of ESPResSo, some obstacles to the
development processes of their package had been overcome by the introduction of
continuous integration practices, and a peer review process for contributions.
These practices decrease development and maintenance times. The developer of
TCLB mentioned that their package had two sets of code, for executing the models
on the CPU and GPU, and that maintenance was decreased by introducing macros, a
practice which became a common part of the development process.

Developers were also asked how documentation fits into their development
process. Several developers noted that developer documentation plays an
important role in familiarizing potential contributors to the software system
architecture. Without the guidance that the documentation provides it would be
unlikely that contributions would pass the peer review process.

None of the software packages whose developers were interviewed have a formal
software development model. The packages all have fairly small development
teams. These teams do accept outside contributors, but generally the teams are
tight-knit, often working at the same institution, although one of the packages
has an international team. The developer of ESPResSo noted that while no formal
model is used, their development model is something similar to a combination of
agile and waterfall development models.

The developers noted similar project management processes. For teams of only a
couple of developers, the addition of new features or major changes are
discussed with the entire team. Projects with more than a couple of developers
have lead developer roles. These lead developers review potential additions to
the software. The software packages use GitHub for managing the project.
Typically there are several development branches as well as the master branch.

>>>>

\section{Introduction} \label{secIntro}

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice} such as
reflective and non-reflective boundaries. They consider the behavior of a
collection of particles as a single unit at the mesoscopic scale, between the
nanoscopic and microscopic scales. LBM solvers predict the positional
probability of a collection of particles moving through a lattice structure
following a two step process: i) streaming, where the particles
move along the lattice via links; and, ii) colliding, where energy and
momentum is transferred among particles that collide \citep{bao2011lattice}. As
an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds numbers.
\wss{Zahra might have a better example?  A journal publication isn't likely
going to accept a figure produced for another paper.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modeling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development to provide insight into
its best practices and to offer guidance for future development.

@@

The full template consists of 108 questions categorized under 9 qualities:
\begin{inparaenum}
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum} 
The questions were designed to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. The measures are grouped under headings for
each quality, and one for summary information
(Figure~\ref{measurement_template_image}).   The summary section provides
general information, such as the software name, number of developers, etc.  We
follow the definitions given by \citet{gewaltig2012quality} for the software
categories.  Public means software intended for public use.  Private means
software aimed only at a specific group, while the concept category is used for
software written simply to demonstrate algorithms or concepts. The three
categories of development models are: open source, where source code is freely
available under an open source license; free-ware, where a binary or executable
is provided for free; and, commercial, where the user must pay for the software
product.

@@

We requested interviews from all 24 packages, except Munubi, since Munubi was
added after our ethics approved study window had expired.  We were able to
recruit 4 developers to participate in our study.  To send out interview
requests we found we found contacts on the projects' website, or code
repository, or publications, or the biographic pages of the teams'
institutions. We send at most two interview request emails to a contact for each
software package.  Meeting will typically be held using on-line meeting
software, like Zoom or Teams.  This facilitates recording and automatic
transcription of the meetings.

@@

\begin{table}
	\begin{center}
		\begin{tabular}{ p{3cm}llllllll}
			\hline
			Name & Dim & Pll & Com & Rflx & MFl & Turb & CGE & OS\\
			\hline
			DL\_MESO (LBE) & 2, 3 & MPI/OMP & Y & Y & Y & Y & Y & W, M, L\\
			ESPResSo & 1, 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & M, L\\
			ESPResSo++ & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			HemeLB & 3 & MPI & Y & Y & Y & Y & Y & L\\
			laboetie & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LatBo.jl & 2, 3 & - & Y & Y & Y & N & Y & L\\
			LB2D-Prime & 2 & MPI & Y & Y & Y & Y & Y & W, L\\
			LB3D & 3 & MPI & N & Y & Y & Y & Y & L\\
			LB3D-Prime & 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			lbmpy & 2, 3 & CUDA & Y & Y & Y & Y & Y & L\\
			lettuce & 2, 3 & CUDA & Y & Y & Y & Y & Y & W, M, L\\
			LIMBES & 2 & OMP & Y & Y & N & N & Y & L\\
			Ludwig & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			LUMA & 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			MechSys & 2, 3 & - & Y & Y & Y & Y & Y & L\\
			MP-LABS & 2, 3 & MPI/OMP & N & Y & Y & N & N & L\\
			OpenLB & 1, 2, 3 & MPI & Y & Y & Y & Y & Y & W, M, L\\
			Palabos & 2, 3 & MPI & Y & Y & Y & Y & Y & W, L\\
			pyLBM & 1, 2, 3 & MPI & Y & Y & N & Y & Y & W, M, L\\
			Sailfish & 2, 3 & CUDA & Y & Y & Y & Y & Y & M, L\\
			SunlightLB & 3 & - & Y & Y & N & N & Y & L\\
			TCLB & 2, 3 & CUDA/MPI & Y & Y & Y & Y & Y & L\\
			waLBerla & 2, 3 & MPI & Y & Y & Y & Y & Y & L\\
			\hline
		\end{tabular}
		\caption{Features of Software Packages (Dim for Dimension (1, 2, 3),
			Pll for Parallel (CUDA, MPI, OpenMP (OMP)), Com for Compressible (Yes or No), Rflx for
			Reflexive Boundary Condition (Yes or No), MFl for Multi-fluid (Yes or
			No), Turb for Turbulent (Yes or No), CGE for Complex Geometries (Yes or No), OS for Operating System (Windows (W), macOS (M), Linux (L)))} \label{tbl_features}
	\end{center}
\end{table}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/correctnessverifiability.pdf}
		\caption{AHP Surface Correctness and Verifiability Score}
		\label{Fig_CorrectnessVerifiability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/reliability_chart.pdf}
		\caption{AHP Surface Reliability Score}
		\label{Fig_Reliability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/robustness_chart.pdf}
		\caption{AHP Surface Robustness Score}
		\label{Fig_Robustness}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/usability_chart.pdf}
		\caption{AHP Surface Usability Score}
		\label{Fig_Usability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/maintainability_chart.pdf}
		\caption{AHP Maintainability Score}
		\label{Fig_Maintainability}
	\end{center}
\end{figure}

@@

We measured the total number of source code files for each project. A larger
number of source files is associated with increased reusability, due to our
assumption that this indicates increased modularization. Some packages have more
features than others.  This is assumed to contribute to reusability, since they have
more source code for potential reuse. The software packages were also reviewed
for the presence of API documentation, which indicates that a software package
was developed with interaction between other software applications in mind.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/reusability_chart.pdf}
		\caption{AHP Reusability Score}
		\label{Fig_Reusabilty}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/understandability_chart.pdf}
		\caption{AHP Understandability Score}
		\label{Fig_Understandability}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=1.0\textwidth]{./figures/visibilitytransparency_chart.pdf}
		\caption{AHP Visibility and Transparency Score}
		\label{Fig_VisibilityTransparency}
	\end{center}
\end{figure}

@@

\begin{figure}[h!]
	\centering
		\includegraphics[width=1.0\textwidth]{./figures/finalscore_chart.pdf}
		\caption{AHP Overall Score}
		\label{Fig_OverallScore}
\end{figure}

@@

, without a need for a rigorous software development model. The developer of
ESPResSo noted that while no formal model is used, their development model is
something similar to a combination of agile and waterfall development models.

Improving the software engineering
education or experience of developers is also an idea that was brought up by
several developers.

@@

Interviews with developers confirmed that these software packages are developed
by domain experts with backgrounds in physics, mathematics and mechanical
engineering. \wss{Could point to surveys that have shown the same thing.}  It
was noted in these interviews that some of the developers do not have formal
software engineering education. \wss{Can point to other studies to support
this.} Some of the development teams include computer scientists. Despite a lack
of visible domain documentation and a resulting lower surface correctness and
verifiability score, it is clear that some of the software packages were
developed by teams with significant domain knowledge.

@@

Several developers alluded to difficulty with testing the correctness of large
numbers of features, and some even manually tested program output, as opposed to
using automated testing. The use of well defined unit testing tools could
decrease the time spent testing some feature.  \wss{link this to
recommendations}

@@

Interviews with developers revealed several usability issues. Some users have
misunderstood the boundaries of LBM and CFD, and have combined or applied
methods that are not physically sound. Sometimes users have applied LBM to
poorly defined or inappropriate fluid dynamics problems. For example, they may
wish to model flow through or around a structure despite having limited
information about the structure or its environment, and having little previous
knowledge of CFD. The users do not realize the limitations of the methods, of
the software, and do not understand the requirements to properly model a problem
with the software. As the developer of TCLB noted, such software packages are
not designed to be used ``out of the box'' in a plug and play fashion, and it
could take months or more to set up the CFD problems correctly. Developers of
some software packages, including ESPResSo, mitigated this by editing the source
code to prevent users from ``combining methods that are not physically sound
together'', and by updating the documentation to better inform users of LBM
limitations, and of the requirements to properly model appropriate problems,
including what algorithms and parameters to use.  \wss{Could point to this for
recommendations related to documentation.  Could point to this for a generative
approach that won't allow inadmissible combinations, or at least includes an
explicit warning.}

Usability concerns were also raised for installation, understanding how to
maneuver the interfaces and understanding how to set up and run models. These
issues are addressed by various user support models, including frequently asked
questions sections on the software package websites, user guides, and hardware
and software requirements specifications.  \wss{Could point to there being room
for improvement here, especially for requirements specification.}

One software package (ESPResSo) changed some of its scripting language to Python
to make it more usable. The developer commented that this was ``the biggest step
in terms of usability over the years'', further commenting that ``most people in
the field know [Python]'' and that ``it's easy to learn''.  \wss{A generative
approach facilitates switching to whatever language is currently popular.}

@@

Although not always visible from the publicly available artifacts, interviews
with developers revealed that most projects have a defined process for accepting
contributions from team members. The packages rarely get contributions from
outside developers, but the process would be similar as for the aforementioned
group.  In general, contributions are made through GitHub, and are then reviewed
and pulled by lead developers, often with consultation with a group of core
developers depending on the organizational model. Continuous integration is part
of the process for some packages. \wss{Recommendation to increase.  Successful
projects use it, a reasonable next step for projects that have not yet
incorporated CI.}

Some developers noted that their software package does not have well defined
contributing guide in the repository, but it might be a good idea to add one in
the near future. They would be happy to see contribution from outside of their
organization, but currently this does not happen.  \wss{A comment was made at
the recent DOE workshop that you get new users through documentation and
tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
for the recommendations part.}

Furthermore, maintainability has been addressed by increasing source code
modularity, reducing duplicate information, and improving abstraction by
developing well defined interfaces. This was noted by the developers of ESPResSo
and pyLBM. Several software packages have had sections of their code base
redeveloped with languages that the developers felt are more understandable and
readable, and that are better supported, such as Python. Data structures have
also been redeveloped and storage has been improved. A developer of pyLBM
mentioned that the geometries and models of their system had been ``decoupled'',
using abstraction and modularization of the source code, to make it ``very easy
to add [new] features''.  \wss{Highlights the advantage of design.  Other
projects can look to these examples.  Maybe cite the Anshu paper about Flash?
Dune?  and point to the advantages of design.}

The developer of TCLB mentioned that their package had two sets of code, for
executing the models on the CPU and GPU, and that maintenance was decreased by
introducing macros, a practice which became a common part of the development
process.

@@

Software packages were not quantitatively measured for modifiability. In this
project we asked developers to comment on modifiability when we interviewed
them. Specifically, we asked if ease of future changes to the system, modules,
and code blocks was considered when designing the software. We also asked if any
measures had been taken to ensure the ease of future changes. All of the
developers that were interviewed noted that the ease of future changes was
considered and that measures to ensure it had been taken, including requiring
the separation of software components in the source code architecture.  \wss{A
good approach for other developers to adopt - part of recommendations.  Trace
back to this comment.}  A high degree of code modularity and abstraction was
noted by developers as a measure to ensure the ease of future changes. This can
be ensured by separating components and hiding information behind well defined
interfaces. The developer of ESPResSo also noted that some of the code base was
transitioned from C to C++, which could ease modifiability of that software
package. The developer of TCLB noted that their software package was designed to
allow for the addition of some LBM features, but changes to major aspects of the
system would be difficult. For example, ``implementing a new model will be an
easy contribution'', but changes to the ``Cartesian mesh...will be a
nightmare''. Furthermore, the package was designed with flexible data structures
and storage in mind.

Some software packages, like Palabos, provide validation benchmarks for their
core fundamental algorithmic ingredients \citep{latt2021palabos}. The stated
intent of these benchmarks is to showcase the validity and usefulness of the
package to stimulate the development of third-user extensions. The Palabos
package identifies as a development framework for modeling problems in various
CFD areas.  \wss{Add validation benchmarks as a ``best practice'' used by
Palabos - others could emmulate.  Goal of this exercise is for people in the
community to learn from each other, and potentially from our advice too.}

Developers were asked to comment on the obstacles in their development process.
The developer of ESPResSo noted that a lot of their source code had been written
with a specific application in mind, and that there is too much coupling between
components. Addressing this issue, possibly by making the design review process
more transparent (Section~\ref{secVisib}), would help with code modifiability
and reusability.  \wss{Traceback for recommendations.}

@@

Software developers noted that they believe users have generally found their
packages to be understandable. \wss{Confusion between understandability and
usability - threat to validity example.}  The interviewed developer of ESPResSo
commented that some users have attempted to run physically incompatible LBM
methods, and the solution was to edit the code to prevent such combinations, as
well as to update the documentation to prevent misunderstanding the methods.
Similarly, a developer of pyLBM noted that some users had issues setting up
parameters for LBM schemes. The solution to this was to update the interface
where these parameters are set, as well as to add functionality to test the
stability of the parameters. A developer of OpenLB noted that some users lack
the background knowledge to easily model fluid dynamics problems using their
software. A frequently asked questions section was added to their package
website to help users find answers to common questions. The package also has
detailed documentation, including guides and usage requirements specification,
to better help users understand the software.  \wss{Peter - the discussion here
is more about user usability than the understandability of the code for
developers.  Did any of the developers discuss developer understandability?  The
confusion between understandability and usability was also a confusion during
the interviews Ao conducted.}

\wss{I moved this from the visibility section, since it is more about
understandability.}  Developers were also asked how documentation fits into their
development process. Several developers noted that developer documentation plays
an important role in familiarizing potential contributors to the software system
architecture. Without the guidance that the documentation provides it would be
unlikely that contributions would pass the peer review process.

@@

The interviewed developer of ESPResSo noted that all major additions to their
package had accompanying changes to artifacts and documentation. They noted that
considerable effort had been put into the documentation. They further commented
that they want to lower the entry barrier for new developers, and because of
that their package has a considerable amount of developer documentation. This
documentation informs developers on how to get started, and orients them to the
artifacts, source code, and system architecture, as well as how the software
package build system works, and how the coupling between the simulation engine
and the interface works.  \wss{Peter - these points about developer
documentation seem like they would fit better under understandability, don't you
think?}  \wss{If ESPresSo scored high (I think it did), we can use that to
further promote the value of documentation.  Also a threat to validity is that
we have implicitly assumed documentation is valuable when measuring quality.}

Developers noted the importance of documentation for both the users and
developers of their software. New features are always added to the
documentation. The developers use documentation to stay up to date on the status
of the software package, and to help expand features, like computational models
or algorithms. This is necessary so that the coding standard for these models is
kept consistent for new developers. \wss{Peter - this also sounds like
discussion that is relevant for understandability.}

The importance of documentation for both users and developers was stressed
throughout the interviews. However, it was noted several times that a lack of
time and funding has a negative affect on the documentation. Most of the
developers are scientific researchers evaluated on the scientific papers that
they produce. Writing and updating documentation is something that is done in
their free time, if that time arises. Sometimes it is a last priority for the
developers. Finding ways to hasten updating documentation would increase the
frequency of such updates and benefit both users and developers.  \wss{Same
thing has come up in papers.  I'll try to find a reference for this.  This is
also something we can point to for recommendations related to code generation.}
\wss{Peter - this also sounds like understandability related discussion?  I
wonder if we should drop Traceability as a quality that we assessed?  There
doesn't seem to be much content specific to traceability.}  The developer of
OpenLB noted the use of documentation generators like Doxygen. It would be
advisable for more projects to use such automatic document generation tools,
since some projects do not do this.  \wss{We can refer back to this for the
recommendations.}

\subsection{Visibility and Transparency} \label{secVisib}

For the developers interviewed, visibility is not currently a priority.
Visibility is not particularly necessary because the teams are fairly small.
Although they accept outside contributors, these are rare and the teams remain
tight-knit.  Most of the teams interviewed have all members at the same
institution, although one team is spread out internationally. \wss{Peter, can
you say which team is international, so that we can add this detail to the
paper?}  For teams of only a few developers, new features or major changes are
discussed with the entire team. Projects with more than a few developers have
lead developer roles. These lead developers review potential additions to the
software.

Although visibility is not a priority, positive steps are being undertaken. For
instance, all interviewed teams used GitHub, with its visibility enhancing
features of branching, issue tracking, insights on contributors, the blame
feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
makes projects even more visible, since contributors can explicitly see the bar
that must be crossed for code to be incorporated into the master branch.
ESPResSo also uses peer review, which improves visibility by explicitly asking
for feedback from other developers.
